---
title: "Human Health and Economic costs of Storm Events Between 1950 and 2011"
author: "ehagan"
output: 
  html_document:
    keep_md: true
---

# Synopsis

This report describes the damage done by storm events to human health and economic activity.


# Loading and Processing the Data

From the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database
contains data for major storms, specific events resulting, and the human and economic costs of those
events. We are working from a copy cached for use with the Coursera Reproducible Research class.

```{r workingDir,echo=FALSE}
# Working from GitHub directory
setwd("D:/GitHub/datasciencecoursera/RepData_PeerAssessment2/")
```

Download the NOAA storm database.
```{r downloadData}
dataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName = "StormData.csv.bz2"

if (! file.exists(fileName)) {
    # Download and check return code. 0 is success, non-zero is error. May also throw exception.
    if (0 != download.file(url=dataUrl, destfile=fileName)) {
        stop("Failed to download data file: " + dataUrl)
    }
}
```

Compute database checksum and verify it matches the known checksum.
```{r checkSum}
library(tools)
checkSum <- md5sum(fileName)
if (checkSum != "df4aa61fff89427db6b7f7b1113b5553") {
    stop("Data md5sum does not match known value")
}
```

Load database if it is not already cached.
```{r readDataFile,cache=TRUE}
# Read csv can automatically handle compressed csv files.
if(!exists("weatherData") || (nrow(weatherData) != 902297) || (ncol(weatherData) != 37))  {
    weatherData <- as.data.frame(read.csv(fileName))
}
```

Compute summary statistics for fatalities and injuries by event type.
```{r cache=TRUE}
fatalitySum  <- aggregate(FATALITIES ~ EVTYPE, data=weatherData, sum)
fatalityMean <- aggregate(FATALITIES ~ EVTYPE, data=weatherData, mean)
injurySum    <- aggregate(INJURIES ~ EVTYPE, data=weatherData, sum)
injuryMean   <- aggregate(INJURIES ~ EVTYPE, data=weatherData, mean)
```

Sort the summary data so that the most damaging events are first and create plots.
```{r}
```

http://www.r-bloggers.com/how-do-i-re-arrange-ordering-a-plot-revisited/
http://rpubs.com/rdpeng/13396
https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf