---
title: "Human Health and Economic Costs of Storm Events Between 1950 and 2011"
author: "ehagan"
output: 
  html_document:
    keep_md: true
---

# Synopsis

The goal of this project is to determine which Storm Event Types are most damaging in terms of human casualty and economic loss. Tornadoes have been, and remain the largest threat to human life from storm activity in the United States. However, in recent years, tornadoes have become comparatively less threatening. Extreme heat, flash flooding and lightning are now signficant contributers. Property damage is mostly caused by water in one form or another along with wind. Crop damage also tends to be caused by water but the most dominant cause is lack of water, i.e. drought.

# Data Processing

From the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database
contains data for major storms, specific events resulting, and the human and economic costs of those
events. Here, I am working from a copy cached for use with the Coursera Reproducible Research class.

```{r workingDir,echo=FALSE}
# Working from GitHub directory
#setwd("D:/GitHub/datasciencecoursera/RepData_PeerAssessment2/")
setwd("/Users/ehagan/github/RepData_PeerAssessment2/")
options(width=800)
```

Download the NOAA storm database.
```{r downloadData}
dataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName = "StormData.csv.bz2"

if (! file.exists(fileName)) {
    # Download and check return code. 0 is success, non-zero is error. May also throw exception.
    if (0 != download.file(url=dataUrl, destfile=fileName, method="curl")) {
        stop("Failed to download data file: " + dataUrl)
    }
}
```

Compute database checksum and verify it matches the known checksum.
```{r checkSum}
library(tools)
checkSum <- md5sum(fileName)
if (checkSum != "df4aa61fff89427db6b7f7b1113b5553") {
    stop("Data md5sum does not match known value")
}
```

Load database if it is not already cached.
```{r readDataFile,cache=TRUE}
# Read csv can automatically handle compressed csv files.
if(!exists("weatherData") || (nrow(weatherData) != 902297) || (ncol(weatherData) != 37))  {
    weatherData <- as.data.frame(read.csv(fileName))
}
```

Compute summary statistics for fatalities and injuries by event type.
The values plotted will only be the top cumulative 80% of casualties.
```{r casualtyData,cache=TRUE}
targetQuantile = 0.9
lastDecade = as.Date("2001/01/01")

fatalitySum  <- aggregate(FATALITIES ~ EVTYPE, data=weatherData, sum)
fatalitySumRecent  <- aggregate(FATALITIES ~ EVTYPE, data=weatherData[which(as.Date(weatherData$BGN_DATE,"%m/%d/%Y %H:%M:%S") > lastDecade),], sum)

fatalitySum  <- fatalitySum[order(-fatalitySum$FATALITIES),]
fatalitySumRecent  <- fatalitySumRecent[order(-fatalitySumRecent$FATALITIES),]

fatalitySum$EVTYPE  <- factor(fatalitySum$EVTYPE, levels=fatalitySum[order(fatalitySum$FATALITIES), "EVTYPE"])
fatalitySumRecent$EVTYPE  <- factor(fatalitySumRecent$EVTYPE, levels=fatalitySumRecent[order(fatalitySumRecent$FATALITIES), "EVTYPE"])

fatalitySum  <- within(fatalitySum, cSum <- cumsum(FATALITIES))
fatalitySumRecent  <- within(fatalitySumRecent, cSum <- cumsum(FATALITIES))

pctfS      <- fatalitySum[fatalitySum$cSum < targetQuantile * sum(fatalitySum$FATALITIES),]
pctfSR     <- fatalitySumRecent[fatalitySumRecent$cSum < targetQuantile * sum(fatalitySumRecent$FATALITIES),]

injurySum    <- aggregate(INJURIES ~ EVTYPE, data=weatherData, sum)
injurySumRecent    <- aggregate(INJURIES ~ EVTYPE, data=weatherData[which(as.Date(weatherData$BGN_DATE,"%m/%d/%Y %H:%M:%S") > lastDecade),], sum)

injurySum    <- injurySum[order(-injurySum$INJURIES),]
injurySumRecent    <- injurySumRecent[order(-injurySumRecent$INJURIES),]

injurySum$EVTYPE  <- factor(injurySum$EVTYPE, levels=injurySum[order(injurySum$INJURIES), "EVTYPE"])
injurySumRecent$EVTYPE  <- factor(injurySumRecent$EVTYPE, levels=injurySumRecent[order(injurySumRecent$INJURIES), "EVTYPE"])

injurySum    <- within(injurySum, cSum <- cumsum(INJURIES))
injurySumRecent    <- within(injurySumRecent, cSum <- cumsum(INJURIES))

pctiS      <- injurySum[injurySum$cSum < targetQuantile * sum(injurySum$INJURIES),]
pctiSR      <- injurySumRecent[injurySumRecent$cSum < targetQuantile * sum(injurySumRecent$INJURIES),]
```

Compute summary statistics for property and crop damage by event type.
The values plotted will only be the top cumulative 90% of casualties.
The database records damage as a value and an "exponent" which is a character
code of dubious reliability. Leaving aside that the standard 32bit Int can resolve monetary
values in excess of 1 Billion dollars, we'll have to tranlate from letters to numbers with 
a mapping. We discard rows where the DMG value is not greater than zero.
```{r economicData,cache=TRUE}
expMapChars = c("h","H","k","K","m","M","b","B")
expMapVals = c(100,100,1000,1000,1000000,1000000,1000000000,1000000000)
expMap = cbind(expMapChars, expMapVals)

# Property damage
propDmg  <- weatherData[which(weatherData$PROPDMG > 0),]
propDmg$PROPDMGEXP <- expMap[match(propDmg$PROPDMGEXP, expMap),2]
propDmg$PROPDMG_COMP <- as.numeric(propDmg$PROPDMG) * as.numeric(propDmg$PROPDMGEXP)
propSum  <- aggregate(PROPDMG_COMP ~ EVTYPE, data=propDmg, sum)
propSum$EVTYPE  <- factor(propSum$EVTYPE, levels=propSum[order(propSum$PROPDMG_COMP), "EVTYPE"])
propSum    <- propSum[order(-propSum$PROPDMG_COMP),]
propSum    <- within(propSum, cSum <- cumsum(PROPDMG_COMP))
pctpdS      <- propSum[propSum$cSum < targetQuantile * sum(propSum$PROPDMG_COMP),]

cropDmg  <- weatherData[which(weatherData$CROPDMG > 0),]
cropDmg$CROPDMGEXP <- expMap[match(cropDmg$CROPDMGEXP, expMap),2]
cropDmg$CROPDMG_COMP <- as.numeric(cropDmg$CROPDMG) * as.numeric(cropDmg$CROPDMGEXP)
cropSum  <- aggregate(CROPDMG_COMP ~ EVTYPE, data=cropDmg, sum)
cropSum$EVTYPE  <- factor(cropSum$EVTYPE, levels=cropSum[order(cropSum$CROPDMG_COMP), "EVTYPE"])
cropSum    <- cropSum[order(-cropSum$CROPDMG_COMP),]
cropSum    <- within(cropSum, cSum <- cumsum(CROPDMG_COMP))
pctcdS      <- cropSum[cropSum$cSum < targetQuantile * sum(cropSum$CROPDMG_COMP),]
```


# Results

Plot total fatalities and injuries by event type. The plots show that tornadoes are consistently the
largest threat to human life and health from single storm event. 
```{r plotHumanCasualties, fig.width=10, fig.height=8}
library(ggplot2)
library(gridExtra)
fatalityPlot = ggplot(pctfS, aes(x=EVTYPE, y=FATALITIES)) + ggtitle("Human fatalities 1950-2011") + geom_bar(stat="identity") + coord_flip()
fatalityRecentPlot = ggplot(pctfSR, aes(x=EVTYPE, y=FATALITIES)) + ggtitle("Human fatalities 2001-2011") + geom_bar(stat="identity") + coord_flip()

injuryPlot   = ggplot(pctiS, aes(x=EVTYPE, y=INJURIES)) + ggtitle("Human injuries 1950-2011") + geom_bar(stat="identity") + coord_flip()
injuryRecentPlot   = ggplot(pctiSR, aes(x=EVTYPE, y=INJURIES)) + ggtitle("Human injuries 2001-2011") + geom_bar(stat="identity") + coord_flip()
grid.arrange(fatalityPlot, injuryPlot, fatalityRecentPlot, injuryRecentPlot, ncol=2, nrow=2, main=sprintf("Human casualties vs. Storm Event Type: Top %d pct.", targetQuantile * 100))
```

Create plots of economic damages. These plots show that flooding and large, powerful storm systems like
Hurricanes and Tornadoes do the most property damage. They also show that drought does the most damage to crops.
```{r economicPlots, fig.width=10, fig.height=6}
propDmgPlot   = ggplot(pctpdS, aes(x=EVTYPE, y=PROPDMG_COMP)) + ggtitle("Property Damage 1950-2011") + geom_bar(stat="identity") + coord_flip()
cropDmgPlot   = ggplot(pctcdS, aes(x=EVTYPE, y=CROPDMG_COMP)) + ggtitle("Crop Damage 1950-2011") + geom_bar(stat="identity") + coord_flip()
grid.arrange(propDmgPlot, cropDmgPlot, nrow=2)
```
